Approximation as Differentiation

### 2025-03-26 notes/sec/grothendieck.tex
Done with g' case, just the indexed family part of f' to grok now.
Correction.
First version.
Clarify triangle identities.
C -> D.

\begin{definition}[Grothendieck construction for $F$]
Suppose a functor $F: D \to \Cat$. The \emph{Grothendieck construction} $\Grothendieck{D}F$ for $F$ is the
category where:
\begin{itemize}
\item objects $X$ are pairs of an object $\idx{X}$ of $D$ and an object $\faml{X}$ of $F(\idx{X})$;
\item morphisms $f: X \to Y$ are morphisms $\idx{f}: \idx{X} \to \idx{Y}$ in $D$ paired with morphisms
$\faml{f}: F(f)(\faml{X}) \to \faml{Y}$ in $F(\idx{Y})$.
\end{itemize}
\end{definition}

\noindent When $F$ is contravariant, a morphism $f: X \to Y$ is a morphism $\idx{f}: \idx{X} \to \idx{Y}$ in
$D$ paired with a morphism $\faml{f}: \faml{X} \to F(f)(\faml{Y})$ in $F(\idx{X})$.

@@

\begin{enumerate}
\item $[X, Y]$ to be the object $(X',Y')$ in $\Fam(C)$ with $X' = \Fam(C)(X,Y)$ and $Y'_f =
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}}$ for every $f: X \to Y$.
\item The family of isomorphisms $\lambda_{X,Y,Z}$ (natural in $X$) sending any $(f, g): X \times Y \to Z$ to
$(f', g'): X \to [Y, Z]$ with:
\begin{itemize}
\item $f'(x) = (f \comp (x, -), \reindex{g}{(x,-)} \comp \inj_{\faml{Y}}):
Y \to Z$ where $\inj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g}{(x,-)} \comp \inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f})$ where $\inj_{\faml{X}_x}: \const_{\idx{Y}}(\faml{X}_x)
\to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\end{itemize}
\item The family of morphisms $\eval_{X,Y} = (f',g'): [X, Y] \times X \to Y$ in $\Fam(C)$ with
\begin{itemize}
\item $f'(f, x) = \idx{f}(x)$
\item $g'_{f, x} = \coprodM{{\eval_{\prod}}_x}{\faml{f}_x}:
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}} \biprod \faml{X}_x \to \faml{Y}_{\idx{f}(x)}$
\end{itemize}
\end{enumerate}

Then:
\begin{enumerate}
\item Suppose $(f,g): X \to [Y,Z]$ and define $(f^\dagger, g^\dagger): X \times Y \to Z$ with:
\begin{itemize}
\item $f^\dagger(x,y) = \idx{f(x)}(y)$
\item $g^\dagger_{x,y} = \coprodM{{\eval_{\prod}}_y \comp g_x}{\faml{f(x)}_y}$
\end{itemize}

To show $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f,g)$, suppose $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f',
g')$ with
\begin{itemize}
\item $f'(x) = (f^\dagger \comp (x, -), \reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{Y}}): Y \to Z$ where
$\inj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f^\dagger})$ where $\inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\end{itemize}
\noindent To show $f' = f : \idx{X} \to \Fam(C)(Y,Z)$, suppose $x \in \idx{X}$. To show $f'(x) = f(x): Y \to
Z$, suppose $y \in \idx{Y}$.
\begin{enumerate}
\item $(f^\dagger \comp (x, -))(y) = f^\dagger(x,y) = \idx{f(x)}(y)$.
\item $(\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{Y}})_y = \reindex{g^\dagger}{(x,-)}_y \comp
{\inj_{\faml{Y}}}_y = g^\dagger_{x,y} \comp {\inj_{\faml{Y}}}_y = \coprodM{{\eval_{\prod}}_y \comp
g_x}{\faml{f}_y} \comp {\inj_{\faml{Y}}}_y = g_x$.
\end{enumerate}

To show $g' = g$, suppose $x \in \idx{X}$. To show $g'_x = g_x$:
\begin{align*}
\small
g'_x
&=
\lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{X}_x}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{g^\dagger_{(x,y)} \comp (\inj_{\faml{X}_x})_y\}_{y \in \idx{Y}}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{\coprodM{{\eval_{\prod}}_y \comp g_x}{\faml{f(x)}_y} \comp (\inj_{\faml{X}_x})_y\}_{y \in \idx{Y}}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{\coprodM{{\eval_{\prod}}_y \comp g_x}{\faml{f(x)}_y}\}_{y \in \idx{Y}})
\end{align*}

\item Suppose $f: X \times Y \to Z$ in $\Fam(C)$ and $(x, y) \in \idx{X} \times \idx{Y}$. The diagram on the
left below commutes by the universal property of the set-indexed product; the commutativity of the diagram on
the right is trivial.

>>>>

\begin{definition}[Grothendieck construction for $F$]
Suppose a functor $F: C \to \Cat$. The \emph{Grothendieck construction} $\Grothendieck{C}F$ for $F$ is the
category where:
\begin{itemize}
\item objects $X$ are pairs of an object $\idx{X}$ of $C$ and an object $\faml{X}$ of $F(\idx{X})$;
\item morphisms $f: X \to Y$ are morphisms $\idx{f}: \idx{X} \to \idx{Y}$ in $C$ paired with morphisms
$\faml{f}: F(f)(\faml{X}) \to \faml{Y}$ in $F(\idx{Y})$.
\end{itemize}
\end{definition}

\noindent When $F$ is contravariant, a morphism $f: X \to Y$ is a morphism $\idx{f}: \idx{X} \to \idx{Y}$ in
$C$ paired with a morphism $\faml{f}: \faml{X} \to F(f)(\faml{Y})$ in $F(\idx{X})$.

@@

\begin{enumerate}
\item $[X, Y]$ to be the object $(X',Y')$ in $\Fam(C)$ with $X' = \Fam(C)(X,Y)$ and $Y'_f =
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}}$ for every $f: X \to Y$.
\item The family of isomorphisms $\lambda_{X,Y,Z}$ (natural in $X$) sending any $(f, g): X \times Y \to Z$ to
$(f', g'): X \to [Y, Z]$ with:
\begin{itemize}
\item $f'(x) = (f \comp (x, -), \reindex{g}{(x,-)} \comp \inj_{\faml{Y}}):
Y \to Z$ where $\inj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g}{(x,-)} \comp \inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f})$ where $\inj_{\faml{X}_x}: \const_{\idx{Y}}(\faml{X}_x)
\to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\end{itemize}
\item The family of morphisms $\eval_{X,Y}: [X, Y] \times X \to Y$ in $\Fam(C)$ with
\begin{itemize}
\item $\idx{\eval_{X,Y}}(f, x) = \idx{f}(x)$
\item $\faml{\eval_{X,Y}}_{f, x} = \coprodM{{\eval_{\prod}}_x}{\faml{f}_x}:
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}} \biprod \faml{X}_x \to \faml{Y}_{\idx{f}(x)}$
\end{itemize}
\end{enumerate}

Then the following triangle identities hold:
\begin{enumerate}
\item $\lambda_{X,Y,Z}(\eval_{Y,Z} \comp (f \tensor \id_Y) = f$. Suppose $(f,g): X \to [Y,Z]$ and define
$(f^\dagger, g^\dagger): X \times Y \to Z$ with:
\begin{itemize}
\item $f^\dagger(x,y) = \idx{f(x)}(y)$
\item $g^\dagger_{x,y} = \coprodM{{\eval_{\prod}}_y \comp g_x}{\faml{f(x)}_y}$
\end{itemize}

To show $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f,g)$, suppose $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f',
g')$. Then:
\begin{itemize}
\item $f'(x) = (f^\dagger \comp (x, -), \reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{Y}}): Y \to Z$ where
$\inj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f^\dagger})$ where $\inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$.
\end{itemize}
\noindent To show $f' = f : \idx{X} \to \Fam(C)(Y,Z)$, suppose $x \in \idx{X}$ and show $f'(x) = f(x): Y \to
Z$. Suppose $y \in \idx{Y}$.
\begin{enumerate}
\item $\idx{f'(x)}(y) = (f^\dagger \comp (x, -))(y) = f^\dagger(x,y) = \idx{f(x)}(y)$.
\item $\faml{f'(x)}_y = (\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{Y}})_y = \reindex{g^\dagger}{(x,-)}_y \comp
{\inj_{\faml{Y}}}_y = g^\dagger_{x,y} \comp {\inj_{\faml{Y}}}_y = \coprodM{{\eval_{\prod}}_y \comp
g_x}{\faml{f(x)}_y} \comp {\inj_{\faml{Y}}}_y = \faml{f(x)}_y$.
\end{enumerate}

To show $g' = g$, suppose $x \in \idx{X}$.
\begin{align*}
\small
g'_x
&=
\lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{X}_x}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{g^\dagger_{(x,y)} \comp (\inj_{\faml{X}_x})_y\}_{y \in \idx{Y}}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{\coprodM{{\eval_{\prod}}_y \comp g_x}{\faml{f(x)}_y} \comp (\inj_{\faml{X}_x})_y\}_{y \in \idx{Y}}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{{\eval_{\prod}}_y \comp g_x\}_{y \in \idx{Y}}) \\
&=
g_x
\end{align*}
\qed

\item $\eval_{Y,Z} \comp (\lambda_{X,Y,Z}(f) \tensor \id_{Y}) = f$. Suppose $f: X \times Y \to Z$ in $\Fam(C)$
and $(x, y) \in \idx{X} \times \idx{Y}$. The diagram on the left below commutes by the universal property of
the set-indexed product; the commutativity of the diagram on the right is trivial.
